---
title: "Lab #0"
author: "Jared Lyon"
date: "January 26, 2022"
output: html_document
---

```{r setup, include=FALSE}
# This is where you set global options (leave as-is)
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1

### Question 1 

This is an R markdown template. Any text written outside of the code chunks will
be treated as regular text (i.e., ignored by the R interpreter). All R commands 
need to be inserted into code chunks such as the one below:

```{r}
# This is the first code chunk demonstrating basic R commands
summary(pressure)
```

The code chunks can be executed individually by clicking on the green **play** icon
to the right. You can add more code chunks by clicking on the **Insert** icon 
in the toolbar of the RStudio editor pane.

### Question 2

You can also embed plots in your R markdown file:

```{r}
# This is the second chunk demonstrating a simple figure
plot(pressure, xlab = "Temperature", ylab  = "Pressure")
```

Clicking on the green **play** icon to the right of the code chunk will display the
figure right below the code. Once you have answered all of the questions,
you can compile the entire document by clicking on the **Knit** icon in the 
toolbar of the RStudio editor pane This will produce an HTML file and display
it in the **Viewer** pane of RStudio.

# begin task stuff below
```{r}
str(pisco)
```
```{r}
# Use functions subset and aggregate to compute the mean of chla_mean for each latitude between 36 and 50 degrees North
pisco.chla.36.50 <- subset(pisco, subset = latitude >= 36 && latitude <= 50, select = c("chla", "latitude"))
head(pisco.chla.36.50)
aggregate(chla ~ latitude, FUN = mean, data = pisco)
```
