---
title: "Project 3"
author: "Jared Lyon"
date: "4/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1
```{r}
# read in data
benguela <- read.csv("http://faraway.neu.edu/biostats/assn3_benguela.csv")
california <- read.csv("http://faraway.neu.edu/biostats/assn3_california.csv")
canary <- read.csv("http://faraway.neu.edu/biostats/assn3_canary.csv")
humboldt <- read.csv("http://faraway.neu.edu/biostats/assn3_humboldt.csv")
str(benguela)
str(california)
str(canary)
str(humboldt)
```

## 1.
```{r}
# add before/after column
california$period <- ifelse(california$year < 2025 , "before", "after")
benguela$period <- ifelse(benguela$year < 2025, "before", "after")
canary$period <- ifelse(canary$year < 2025, "before", "after")
humboldt$period <- ifelse(humboldt$year < 2025, "before", "after")
```

## 2.
```{r}
# Multimodel mean for California
california.mean <- data.frame(year = california$year, period = california$period,
upwelling = rowMeans(california[1:22]))
# Multimodel mean for Canary
canary.mean <- data.frame(year = canary$year, period = canary$period,
upwelling = rowMeans(canary[1:22]))
# Multimodel mean for Humboldt
humboldt.mean <- data.frame(year = humboldt$year, period = humboldt$period,
upwelling = rowMeans(humboldt[1:22]))
# Multimodel mean for Benguela
benguela.mean <- data.frame(year = benguela$year, period = benguela$period,
upwelling = rowMeans(benguela[1:22]))

california.mean
canary.mean
humboldt.mean
benguela.mean
```

## 3.
In order to test for normality, a Shapiro-Wilk test will be applied to the datapoints of each set. Eight tests total are needed, since each of the four datasets has two periods. In all eight tests, the hypotheses are:
H0: The sample population is normally distributed.
HA: The sample population is not normally distributed.
```{r}
# shapiro test for before periods
shapiro.test(as.matrix(subset(california.mean, period == "before", select = upwelling)))
shapiro.test(as.matrix(subset(canary.mean, period == "before", select = upwelling)))
shapiro.test(as.matrix(subset(humboldt.mean, period == "before", select = upwelling)))
shapiro.test(as.matrix(subset(benguela.mean, period == "before", select = upwelling)))

# shapiro test for after periods
shapiro.test(as.matrix(subset(california.mean, period == "after", select = upwelling)))
shapiro.test(as.matrix(subset(canary.mean, period == "after", select = upwelling)))
shapiro.test(as.matrix(subset(humboldt.mean, period == "after", select = upwelling)))
shapiro.test(as.matrix(subset(benguela.mean, period == "after", select = upwelling)))
```

The p-values for all 'before' periods within the California, Benguela, Canary, and Humboldt models are >0.05, meaning the null hypothesis can be accepted and that these datasets are normally distributed. However, the p-values for the 'after' datasets for Canary and Humboldt are <0.05, meaning that we can reject the null hypothesis for these two sets and that they are not normally distributed. Lastly, the p-values of the 'after' datasets for California and Benguela are >0.05, thus we can accept the null hypothesis for these groups and assert that they are normally distributed.

Because two of our eight tests showed an absence of normality, this means we must resort to non-parametric tests. Although non-parametric tests can be applied to more situations than their parametric counterparts since they do not rely on any distribution types, they have less statistical power due to a lack of assumptions that they make about the data.

## 4.
Since we must resort to non-parametric tests, a Mann-Whitney-Wilcox Test is the most applicable in order to compare the group means between each demographics 'before' and 'after' subsets. On the other hand, a t-test is rejected since the t-test assumes normality and is thus a parametric test. The test hypotheses are:
H0: The sample means are equal.
HA: The sample means are not equal.
```{r}
#conduct wilcox test on each of the four pairs
# Canary
model.upwelling.canary <- wilcox.test(as.matrix(subset(canary.mean,
period == "before", select = upwelling)), as.matrix(subset(canary.mean,
period == "after", select = upwelling)))
model.upwelling.canary

# California
model.upwelling.california <- wilcox.test(as.matrix(subset(california.mean,
period == "before", select = upwelling)), as.matrix(subset(california.mean,
period == "after", select = upwelling)))
model.upwelling.california

# Humboldt
model.upwelling.humboldt <- wilcox.test(as.matrix(subset(humboldt.mean,
period == "before", select = upwelling)), as.matrix(subset(humboldt.mean,
period == "after", select = upwelling)))
model.upwelling.humboldt

# Benguela
model.upwelling.benguela <- wilcox.test(as.matrix(subset(benguela.mean,
period == "before", select = upwelling)), as.matrix(subset(benguela.mean,
period == "after", select = upwelling)))
model.upwelling.benguela
```

The results of all four tests result in p-values lower than 0.05, meaning we can reject the null hypothesis and assert that none of the 'before' vs. 'after' demographic pairs have equal means.

## 5.
```{r}
# Compute group means and stderrs for California
mean.upwelling.california <- aggregate(upwelling ~ period, FUN = mean,
data = california.mean)
sderr.upwelling.california <- aggregate(upwelling ~ period, FUN = function(x) sd(x)/sqrt(length(x)),
data = california.mean)

# Compute group means and stderrs for Canary
mean.upwelling.canary <- aggregate(upwelling ~ period, FUN = mean, data = canary.mean)
sderr.upwelling.canary <- aggregate(upwelling ~ period, FUN = function(x) sd(x)/sqrt(length(x)),
data = canary.mean)

# Compute group means and stderrs for Humboldt
mean.upwelling.humboldt <- aggregate(upwelling ~ period, FUN = mean,
data = humboldt.mean)
sderr.upwelling.humboldt <- aggregate(upwelling ~ period, FUN = function(x) sd(x)/sqrt(length(x)),
data = humboldt.mean)

# Compute group means and stderrs for Benguela
mean.upwelling.benguela <- aggregate(upwelling ~ period, FUN = mean,
data = benguela.mean)
sderr.upwelling.benguela <- aggregate(upwelling ~ period, FUN = function(x) sd(x)/sqrt(length(x)),
data = benguela.mean)
```

## 6.
```{r}
# Combine data into a single means dataframe
upwelling.mean <- data.frame(region = c("Canary", "California",
"Benguela", "Humboldt"), after = c(mean.upwelling.canary$upwelling[2],
mean.upwelling.california$upwelling[2], mean.upwelling.benguela$upwelling[2],
mean.upwelling.humboldt$upwelling[2]), before = c(mean.upwelling.canary$upwelling[1],
mean.upwelling.california$upwelling[1], mean.upwelling.benguela$upwelling[1],
mean.upwelling.humboldt$upwelling[1]))
upwelling.mean

# Combine data into a single stderrs dataframe
upwelling.sderr <- data.frame(region = c("Canary", "California",
"Benguela", "Humboldt"), after = c(sderr.upwelling.canary$upwelling[2],
sderr.upwelling.california$upwelling[2], sderr.upwelling.benguela$upwelling[2],
sderr.upwelling.humboldt$upwelling[2]), before = c(sderr.upwelling.canary$upwelling[1],
sderr.upwelling.california$upwelling[1], sderr.upwelling.benguela$upwelling[1],
sderr.upwelling.humboldt$upwelling[1]))
upwelling.sderr

## plot the results
bp <- barplot(t(as.matrix(upwelling.mean[, 2:3])), beside = TRUE, ylim = c(0, 1), xlab = "Region", names = upwelling.mean$region, ylab = "Upwelling", main = "Before vs. After Upwelling Comparisons for EBCS Regions", legend = TRUE)
arrows(x0 = bp, y0 = t(as.matrix(upwelling.mean[, 2:3])) - t(as.matrix(upwelling.sderr[, 2:3])), y1 = t(as.matrix(upwelling.mean[, 2:3])) + t(as.matrix(upwelling.sderr[, 2:3])), code = 3, angle = 90, length = 0.1)
legend
```

## 7.
Based on the results of the prior statistical tests and the figure produced, we can conclude that upwelling statistically decreases across all four models. One particular item to note within the figure is that none of the standard error bars overlap, further indicating that there is a statistically significant decrease in upwelling over time.

# Problem 2
## 1.
The first possible test to compare sample variances is the parametric F-test, which assumes independent observations and normal distributions. The second possible test is the parametric Levene Test, which also compares sample variances; this test also assumes independent observations, but can be applied to quasi-normal distributions.

## 2.
Given that four tests will be conducted in order to compare the variances between two groups ('before' vs. 'after'), the most applicable test to examine the variances of the groups is the Levene Test since not all of our groups are normally distributed. The hypotheses for this test are as follows:
H0: The variances of the groups are equal.
HA: The variances of the groups are not equal.
```{r}
# install levene test
library(car)
```

```{r}
# conduct Levenetest on each of the four pairs
# Canary
var.upwelling.canary <- leveneTest(upwelling ~ period, canary.mean)
var.upwelling.canary

# California
var.upwelling.california <- leveneTest(upwelling ~ period, california.mean)
var.upwelling.california

# Humboldt
var.upwelling.humboldt <- leveneTest(upwelling ~ period, humboldt.mean)
var.upwelling.humboldt

# Benguela
var.upwelling.benguela <- leveneTest(upwelling ~ period, benguela.mean)
var.upwelling.benguela
```

Based on the p-values of our test results, we can conclude that the variances differ at the Benguela site since the significance code indicates a p-value that is >0.05. For all other sites, however, the p-values do not exceed 0.05, meaning we can reject the null hypothesis at the California, Humboldt, and Canary sites and assert that the variances statistically differ in a significant way.

## 3.
```{r}
# Compute variance for California
var.upwelling.california <- aggregate(upwelling ~ period, FUN = var,
  data = california.mean)
# Compute variance for Canary
var.upwelling.canary <- aggregate(upwelling ~ period, FUN = var,
  data = canary.mean)
# Compute variance for Humboldt
var.upwelling.humboldt <- aggregate(upwelling ~ period, FUN = var,
  data = humboldt.mean)
# Compute variance for Benguela
var.upwelling.benguela <- aggregate(upwelling ~ period, FUN = var,
  data = benguela.mean)
# Combine data
var.upwelling <- data.frame(region = c("Canary", "California",
  "Benguela", "Humboldt"), after = c(var.upwelling.canary$upwelling[2],
  var.upwelling.california$upwelling[2], var.upwelling.benguela$upwelling[2],
  var.upwelling.humboldt$upwelling[2]), before = c(var.upwelling.canary$upwelling[1],
  var.upwelling.california$upwelling[1], var.upwelling.benguela$upwelling[1],
  var.upwelling.humboldt$upwelling[1]))
var.upwelling

# plot the results
bp <- barplot(t(as.matrix(var.upwelling[, 2:3])), beside = TRUE, ylim = c(0, 0.005), xlab = "Region", names = var.upwelling$region, ylab = "Variance", main = "Before vs. After Upwelling Variance for EBCS Regions", legend = TRUE)
```

## 4.
Upwelling means decreased over time across all four regional groups; however, variance increased over time within the Benguela and California, whereas variance decreased over time in Canary and Humboldt. Because variance increased over time in California and Benguela, this means the robustness and consistency of the upwelling means decreased over time in these regions, whereas lowering variance in Humboldt and Canary raises the robustness and consistency of the upwelling means over time.

# Problem 3
## 1.
Since each EBCS will be examined individually, our explanatory variable is time (year) and the response variable is upwelling. Due to the singular nature of both variables, an analysis utilizing simple linear regression may be applied.

## 2.
The assumptions of the test are that the variables are linearly related, the residuals are independent, the residuals have identical homoscedacities, and that the residuals are normally distributed. All of the assumptions are met with the exception of normality specifically within Canary and Humboldt, wherein we know that the 'after' period does not contain normally distributed data; thus, we cannot extrapolate such that the residuals in these data sets will be normally distributed.
H0: There is no relationship between upwelling and year.
HA: There is a relationship between upwelling and year.
```{r}
# run simple regression tests on each region
benguela.lm = lm(upwelling ~ year, benguela.mean)
summary(benguela.lm)
california.lm = lm(upwelling ~ year, california.mean)
summary(california.lm)
humboldt.lm = lm(upwelling ~ year, humboldt.mean)
summary(humboldt.lm)
canary.lm = lm(upwelling ~ year, canary.mean)
summary(canary.lm)
```

## 3.
```{r}
par(mfrow = c(1, 1))
plot(upwelling ~ year, data = benguela.mean, ylim = c(0.2, 1.1), ylab = expression(paste("Upwelling (",
  m^2/s, ")")), xlab = "Year", col = "blue")
points(upwelling ~ year, data = canary.mean, col = "red")
points(upwelling ~ year, data = california.mean, col = "green")
points(upwelling ~ year, data = humboldt.mean, col = "black")
# Add trend lines from your analysis. Insert the model
# name into the XXXX fields
abline(humboldt.lm, col = "black")
abline(canary.lm, col = "red")
abline(benguela.lm, col = "blue")
abline(california.lm, col = "green")
legend(x = "topleft", lty = 1, col = c("blue", "red", "green",
  "black"), legend = c("Benguela", "Canary", "California",
  "Humboldt"))
```

The figure shows that upwelling increases over time in Benguela, Canary, and Humboldt, whereas it remained steady in California.