---
title: "Project 1"
author: "Jared Lyon"
output: html_document
---

### Problem 1
## 1.
```{r}
# Probability of detecting swine flu given that it was contracted
p.D.F2 <- 0.7
# False detection rate
p.D.opF2 <- 0.1
# Swine flu prevalence
p.F2 <- 0.6
# Calculate probability of detection
p.D <- (p.D.F2 * p.F2) + (p.D.opF2 * (1 - p.F2))

# Calculate the probability of having the swine flu, given detection
p.F2.D <- (p.D.F2 * p.F2)/p.D
p.F2.D
```
The probability of having the swine flu given detection is 0.9130

## 2.
```{r}
p.D.F2 <- 0.7
# Assign four equally spaced prevalance values
p.F2 <- seq(from = 0.2, to = 0.8, by = ((0.8 - 0.2)/3))
# Assign 100 equally spaced false detection rates
p.D.opF2 <- seq(from = 0.01, to = 0.5, by = ((0.5 - 0.01)/99))
# Create matrix
p.F2.D <- matrix(nrow = length(p.F2), ncol = length(p.D.opF2))

# Nested loop to fill matrix
for (i in 1:NROW(p.F2.D)) {
  for (j in 1:NCOL(p.F2.D)) {
    # Calculate probability of having the flu given detection at each point in the above sequences
    p.F2.D[i, j] <- ((p.D.F2 * p.F2[i])/((p.D.F2 * p.F2[i]) + (p.D.opF2[j] * (1 - p.F2[i]))))
  }
}
p.F2.D
```

## 3.
```{r}
# Plot the matrix generated in question 2
# Plot the first row of p.F2.D
plot(p.D.opF2, p.F2.D[1, ], xlab = "False Detection Rate", ylab = "Probability of Having the Flu Given Detection", col = "red", abline(b = 0, a = 0.5, lty = "dotted"))
# Plot the second row of p.F2.D
par(new = TRUE)
plot(p.D.opF2, p.F2.D[2, ], xlab = "False Detection Rate", ylab = "Probability of Having the Flu Given Detection", col = "blue", axes = F)
# Plot the third row of p.F2.D
par(new = TRUE)
plot(p.D.opF2, p.F2.D[3, ], xlab = "False Detection Rate", ylab = "Probability of Having the Flu Given Detection", col = "green", axes = F)
# Plot the fourth row of p.F2.D
par(new = TRUE)
plot(p.D.opF2, p.F2.D[4, ], xlab = "False Detection Rate", ylab = "Probability of Having the Flu Given Detection", col = "orange", axes = F)
# Add a legend to the graph
legend(x = "topright", legend = c("0.2 Disease Prevalence Rate", "0.4 Disease Prevalence Rate", "0.6 Disease Prevalence Rate", "0.8 Disease Prevalence Rate"), col = c("red", "blue", "green", "orange"), pch = 1)
```
## 4.
The dashed line represents a 0.5 probability of having the flu given detection. Any above this line would suggest a testing protocol that is more accurate than not. Overall the figure suggests that the probability of having swine flu given detection lowers as the false detection rate rises; thus, ideal tests have the lowest false detection rate and thus yield the highest probability of detecting true positives. A lower prevalence rate exacerbates the effects of the false detection rate; notice how the downwards curve of the matrix rows increases as the prevalence rate lowers, thus creating an illustration that would suggest the test with a 0.3 false detection rate when disease prevalcne is 0.8 would still be roughly as accurate as a test with a 0.2 false detection rate when disease prevalence is 0.2.